A comprehensive framework to capture the arcana of neuroimaging analysis

### 2019-03-15 manuscript/manuscript.tex
tidied up methods reworkings

At the heart of each Study class is the data specification table, which
specifies both the input and output data of the analysis, and all the
stages in between. There is a one-to-one relationship between entries in
the data specification table and data that is stored in the repository
or will be stored if and when it is derived. Which intermediate
derivatives to include in the data specification table, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the table.

@@

For Study classes that correspond to a known type in the BIDS
standard, a dictionary mapping their data specifications to default
\emph{BidsInput} or \emph{BidsAssocInput} selectors can be specified in the
\emph{default\_bids\_inputs} class attribute. BidsInput
specifies a primary scan in the BIDS standard using its \emph{type}, \emph{modality},
\emph{format} and optionally the \emph{task} it belongs to. BidsAssocInput is used to select
associated files, such as field maps and diffusion encoding matrices in reference
to a primary scan. Default Bids selectors are typically not provided with
a value for the \emph{task} keyword, which can be set at instantiation of
the Study class with the \emph{bids\_task} keyword argument to enable the design of Study
classes that are applicable to any BIDS task.

@@

The syntax for pipeline construction is inspired by a proposed form for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is given a name,
which must be unique amongst the pipelines constructed by the Study. The methods implemented
by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

@@

Node interfaces are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input interface, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the data specification table. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the data specification table
and the format the input is expected in (i.e. a FileFormat for Fileset specifications or Python datatype for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output interface and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output interface (output connections are implied by input connections to other nodes).

If the expected format of a pipeline input or generated format of a pipeline output does not match that
of the corresponding study input or data specification, then a conversion node is implicitly connected
to the pipeline by the framework to perform the required conversion. Pipeline inputs and outputs of
that are conditional on parameters or inputs provided to the study can be specified outside of the
\emph{add} method using the \emph{connect\_input} or
\emph{connect\_output} methods, respectively. Conditional connections between
nodes can be specified via the \emph{connect} method of the pipeline (which simply calls the
method of the same name in the underlying Nipype workflow).

@@

Instead of being set directly, the data and parameter specifications are set
by the metaclass of the Study (i.e. StudyMetaClass),
in order to combine them with corresponding specifications in the class' bases. The combined
data and parameter specifications are constructed by visiting the class'
bases in reverse method resolution order (MRO) and adding specifications
from their \emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

@@

In its most basic form, a ``basic'' repository, is just a file system
directory containing the data to be processed for a single subject.
For multi-subject studies, the root directory should contain separate
subdirectories for each subject in the study, with subject IDs taken
from the subdirectory names. If data was acquired for each subject
over multiple visits, than an additional layer of nested subdirectories
is included in the subject subdirectories, with the visit IDs taken from
the subdirectory names. Note that the directory
repository is similar to the BIDS format, however, there are no naming
conventions in the directory repository, which enables prototyping and
testing of analyses on loosely structured data.

Derivatives are stored by their specification name in a study-specific
namespaces to avoid clashes with separate analyses. In directory
repositories this namespace is a subdirectory named after the study
nested within each visit subdirectory if present or subject subdirectory
otherwise. In BIDS repositories, the namespace is a subdirectory of the
\emph{derivatives} directory, again named after the study. In XNAT
repositories, derivatives for each session are stored in separate
\emph{MrSession} objects alongside the primary session underneath its
\emph{Subject}, and are named
\emph{\textless{}primary-session-name\textgreater{}\_\textless{}study-name\textgreater{}}
(Table \ref{tbl:repo-structure}).

@@

When using BIDS repositories, the selection stage is typically already
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided to override the default
selections is required.

@@

Processor modules control how pipelines generated by a Study
are executed. There currently three Processor modules implemented in
Arcana: \emph{SingleProc}, \emph{MultiProc},
\emph{SlurmProc}, which wrap the LinearPlugin, MultiProcPlugin and SlurmPlugin
Nipype execution plugins. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
trivial.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of: parameter values used by the pipeline,
software versions used by the pipeline, a graph representation of the
underlying Nipype workflow, checksums of inputs, version of Arcana used,
version of Nipype used, subject and visit IDs used by study.

For subsequent analyses, changes with respect to any of the stored values, with
the exception of subject and visit IDs, will be flagged as a mismatch.
Subject or visit IDs only flag a mismatch if any of the inputs, or
inputs of upstream pipelines, are per-visit or per-subject,
respectively, or per-study. How provenance mismatches are handled by the
study is determined by the \emph{reprocess} flag, which is passed to the
Study on initialisation (Table \ref{tbl:reprocess-flag}).

@@

Additionally, a time limit can be passed to the
\emph{reprocess\_time\_limit}, and if the estimated time to reprocess
all the required sessions is less than the time limit they will be
reprocessed and otherwise an exception will be raised.

>>>>

At the heart of each Study class is the data specification table, which
is stored in the \emph{\_data\_specs} class attribute and specifies the
input and output data of the analysis, and all
stages in between. There is a one-to-one relationship between entries in
the data specification table and data that are stored in the repository,
or will be stored if and when they are derived. Which intermediate
derivatives to include in the data specification table, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the table.

@@

For Study classes that correspond to a known type in the BIDS
standard, a dictionary mapping their data specifications to default
\emph{BidsInput} or \emph{BidsAssocInput} selectors can be specified in the
\emph{default\_bids\_inputs} class attribute. BidsInput
specifies a primary scan in the BIDS standard using its \emph{type}, \emph{modality},
\emph{format} and optionally the \emph{task} it belongs to. BidsAssocInput is used to select
associated files, such as field maps and diffusion encoding matrices relative
to a primary scan. Default BIDS selectors are typically not provided with
a value for the \emph{task} keyword, which can be set at instantiation of
the Study class with the \emph{bids\_task} keyword argument to enable the design of Study
classes that are applicable to any BIDS task.

@@

The syntax for pipeline construction is inspired by changes proposed for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is given a name,
which must be unique amongst the pipelines constructed by the Study. The methods implemented
by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

@@

Node interfaces are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input interface, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the data specification table. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the data specification table
and the format the input is expected in (i.e. a FileFormat for Fileset specifications or Python datatype for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output interface and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output interface (output connections to other nodes are implied by input connections of other nodes).

If the expected format of a pipeline input or generated format of a pipeline output does not match that
of the corresponding study input or data specification, then a conversion node is implicitly connected
to the pipeline by the framework to perform the required conversion. Pipeline inputs and outputs
that are conditional on parameters or inputs provided to the study can be specified outside of the
\emph{add} method using the \emph{connect\_input} or
\emph{connect\_output} methods, respectively. Conditional connections between
nodes can be specified via the \emph{connect} method of the pipeline (which simply calls the
method of the same name in the underlying Nipype workflow).

@@

Instead of being set directly, the data and parameter specifications are set
by the metaclass of the Study (i.e. StudyMetaClass)
in order to combine them with corresponding specifications in the class' bases. The combined
data and parameter specifications are constructed by visiting the class'
bases in reverse method resolution order (MRO) and adding specifications
from their \emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

@@

In its most basic form, a ``basic'' repository, is just a file system
directory containing the data to be processed for a single subject.
For multi-subject studies, the root directory should contain separate
subdirectories for each subject in the study, with subject IDs taken
from the subdirectory names. If data was acquired for each subject
over multiple visits, than an additional layer of nested subdirectories
is included in the subject subdirectories, with the visit IDs taken from
the subdirectory names. Note that the basic
repository is similar to the BIDS format, however, there are no naming
conventions in the basic repository, which enables prototyping and
testing of analyses on loosely structured data.

Derivatives are stored by their specification name in a study-specific
namespaces to avoid clashes with separate analyses. In basic
repositories this namespace is a subdirectory named after the study
nested within the lowest layer of the data tree. In BIDS repositories, the namespace is a subdirectory of the
\emph{derivatives} directory, again named after the study. In XNAT
repositories, derivatives for each session are stored in separate
\emph{MrSession} objects alongside the primary session underneath its
\emph{Subject}, and are named
\emph{\textless{}primary-session-name\textgreater{}\_\textless{}study-name\textgreater{}}
(Table \ref{tbl:repo-structure}).

@@

When using BIDS repositories, the selection stage is typically already
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided to override the default
selections if required.

@@

Processor modules control how pipelines generated by a Study
are executed. There currently three Processor modules implemented in
Arcana: \emph{SingleProc}, \emph{MultiProc},
\emph{SlurmProc}, which wrap the LinearPlugin, MultiProcPlugin and SlurmPlugin
Nipype execution plugins, respectively. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
trivial.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of parameter values used by the pipeline,
software versions used by the pipeline, a graph representation of the
underlying Nipype workflow, checksums of inputs, version of Arcana used,
version of Nipype used, subject and visit IDs used by study.

For subsequent analyses, changes with respect to any of the stored values, with
the exception of subject and visit IDs, will be flagged as a mismatch by default.
However, changes can be ignored by providing the path to the attribute to ignore
to the \emph{prov\_ignore} keyword argument at initialisation of the Processor.
Subject or visit IDs only flag a mismatch if any of the inputs, or
inputs of upstream pipelines, are of per-visit or per-subject frequency,
respectively, or per-study frequency. How provenance mismatches are handled by the
study is determined by the \emph{reprocess} flag, which is passed to the
Study on initialisation (Table \ref{tbl:reprocess-flag}).

@@
